# *DBI* objects

*rocker* class encapsulates the *DBI* objects driver, connection and result.
If required, these objects can be directly used with *DBI* functions.
__However, it is recommended to use this option with care!__ Direct usage of *DBI* functions, may disrupt proper function of *rocker* class. Many *DBI* functions are implemented in *rocker* class. Whenever possible, use the *rocker* class functions.

Prepare object

```{r}
db <- rocker::newDB() # New database handling object
db$.drv # Empty driver
db$.con # Empty connection
db$.res # Empty result
```

## DBIDriver-class

```{r}
db$setupSQLite() # Setup SQLite database
db$.drv # 'DBI' DBIDriver-class
```

```{r}
db$getInfoDrv() # 'rocker' class function
```

```{r}
DBI::dbGetInfo(db$.drv) # Direct usage of 'DBI' function on 'rocker' class
```

```{r}
RSQLite::dbGetInfo(db$.drv) # Direct usage of driver package, 'RSQLite', function on 'rocker' class
```

## DBIConnection-class

```{r}
db$connect() # Open connection
db$.con # 'DBI' DBIConnection-class
```

```{r}
db$getInfoCon() # 'rocker' class function
```

```{r}
DBI::dbGetInfo(db$.con) # Direct usage of 'DBI' function on 'rocker' class
```

```{r}
RSQLite::dbGetInfo(db$.con) # Direct usage of driver package, 'RSQLite', function on 'rocker' class
```

Prepare table

```{r}
db$writeTable("mtcars", mtcars) # Create table for testing
```

## DBIResult-class

```{r}
db$sendQuery("SELECT * FROM mtcars;") # Send query
db$.res # 'DBI' DBIResult-class
```

```{r}
db$getInfoRes() # 'rocker' class function
```

```{r}
DBI::dbGetInfo(db$.res) # Direct usage of 'DBI' function on 'rocker' class
```

```{r}
RSQLite::dbGetInfo(db$.res) # Direct usage of driver package, 'RSQLite', function on 'rocker' class
```

Clean up

```{r}
db$clearResult() # Clean up result
db$.res # Empty result
db$disconnect() # Close connection
db$.con # Empty connection
db$unloadDriver() # Reset database handling object
db$.drv # Empty driver
```

# *DBI* functions in *rocker*

Generally, *rocker* function names are related to *DBI* function names. In *rocker* functions, the leading __db__ is removed.

In *DBI* most functions need to be supplied with a driver ^(drv)^, connection ^(conn)^ or result ^(res)^ object. In *rocker*, functions automatically access the corresponding objects ^(.drv,^ ^.con^ ^and^ ^.res)^ stored in the class. 

__*DBI* example__

```{r}
drv <- RSQLite::SQLite() # SQLite driver
DBI::dbCanConnect( # Test parameter
  drv = drv,
  dbname = ":memory:"
)
con <- DBI::dbConnect( # Open connection
  drv = drv,
  dbname = ":memory:"
)
DBI::dbWriteTable(con, "mtcars", mtcars) # Create table for testing
res <- DBI::dbSendQuery(con, "SELECT * FROM mtcars;") # Send query
output <- DBI::dbFetch(res) # Fetch result
DBI::dbClearResult(res) # Clean up result
DBI::dbDisconnect(con) # Close connection
DBI::dbUnloadDriver(drv) # Unload driver
```

__*rocker* example__

```{r}
db <- rocker::newDB(verbose = FALSE) # New database handling object
db$setupDriver( # Setup SQLite database
  drv = RSQLite::SQLite(),
  dbname = ":memory:"
)
db$canConnect() # Test parameter
db$connect() # Open connection
db$writeTable("mtcars", mtcars) # Create table for testing
db$sendQuery("SELECT * FROM mtcars;") # Send query
output <- db$fetch() # Fetch result
db$clearResult() # Clean up result
db$disconnect() # Close connection
db$unloadDriver() # Reset database handling object
```

## List of functions

| *rocker* function | Corresponding *DBI* function | *DBI* object used | Comment |
| --- | --- | --- | --- |
| initialize() | *none* | *none* | |
| print() | *none* | *none* | |
| setupDriver() | *none* | *driver from appropriate package* | Usually, parameters provided to dbConnect() in *DBI* are provided to setupDriver() in *rocker*  |
| setupPostgreSQL() | *none* | *none* | RPostgres::Postgres() is used with *rocker* function setupDriver() |
| setupMariaDB() | *none* | *none* | RMariaDB::MariaDB() is used with *rocker* function setupDriver() |
| setupSQLite() | *none* | *none* | RSQLite::SQLite() is used with *rocker* function setupDriver() |
| unloadDriver() | dbUnloadDriver() | driver | |
| canConnect() | dbCanConnect() | driver | Usually, parameters provided to dbCanConnect() in *DBI* are provided to setupDriver() in *rocker* |
| connect() | dbConnect() | driver | Usually, parameters provided to dbConnect() in *DBI* are provided to setupDriver() in *rocker*  |
| disconnect() | dbDisconnect() | connection | |
| sendQuery() | dbSendQuery() | connection | |
| getQuery() | Is __not__ using dbGetQuery(), but has the same function | connection | Especially, combination of *rocker* functions sendQuery(), fetch() and clearResult()  |
| sendStatement() | dbSendStatement() | connection | |
| execute() | Is __not__ using dbExecute(), but has the same function | connection | Especially, combination of *rocker* functions sendStatement() and clearResult() |
| fetch() | dbFetch() | result | |
| hasCompleted() | dbHasCompleted() | result | |
| getRowsAffected() | dbGetRowsAffected() | result | |
| getRowCount() | dbGetRowCount() | result | |
| columnInfo() | dbColumnInfo() | result | |
| getStatement() | dbGetStatement() | result | |
| clearResult() | dbClearResult() | result | |
| begin() | dbBegin() | connection | |
| commit() | dbCommit() | connection | |
| rollback() | dbRollback() | connection | |
| getInfoDrv() | dbGetInfo() | driver | |
| getInfoCon() | dbGetInfo() | connection | |
| getInfoRes() | dbGetInfo() | result | |
| isValidDrv() | dbIsValid() | driver | |
| isValidCon() | dbIsValid() | connection | |
| isValidRes() | dbIsValid() | result | |
| createTable() | dbCreateTable() | connection | |
| appendTable() | dbAppendTable() | connection | |
| writeTable() | dbWriteTable | connection | |
| readTable() | dbReadTable | connection | |
| removeTable() | dbRemoveTable() | connection | |
| existsTable() | dbExistsTable() | connection | |
| listFields() | dbListFields() | connection | |
| listObjects() | dbListObjects() | connection | |
| listTables() | dbListTables() | connection | |
