# New *rocker* class object

Create new *rocker* database handling object

Option 1

```{r}
db <- rocker::newDB() # New database handling object
```

Option 2

```{r}
db <- rocker::rocker$new() # New database handling object
```

__Terminal output__

Controlling terminal output

```{r}
db <- rocker::newDB(verbose = TRUE) # New database handling object
db$setupPostgreSQL()
db$unloadDriver()
```

```{r}
db$verbose <- FALSE # Terminal output off
db$setupPostgreSQL()
db$unloadDriver()
```

```{r}
db$verbose <- TRUE # Terminal output on (default)
db$setupPostgreSQL()
db$unloadDriver()
```

Structure of terminal output

```
Dctr | Driver load RSQLite
D                          = Driver     (D = loaded,    d = not set)
 c                         = Connection (C = opened,    c = closed)
  t                        = Transation (T = active,    t = no tranastion)
   r                       = Result     (R = available, r = no result)
       Driver load RSQLite = Message text

```

__Optional object ID__

Optionally, rocker object can be labeled with an ID. This can be helpful in case terminal output of multiple rocker objects need to be distinguished.

```{r}
db1 <- rocker::newDB(id = "myDB 1") # New database handling object with ID
db2 <- rocker::newDB(id = "myDB 2") # New database handling object with ID
db1$setupPostgreSQL()
db2$setupMariaDB()
db1$unloadDriver()
db2$unloadDriver()
```

```{r}
db1$id <- NULL # Remove ID
db1$setupSQLite()
db1$unloadDriver()
```

```{r}
db1$id <- "newID 1" # Add new ID
db1$setupSQLite()
db1$unloadDriver()
```

__Object properties__

Object properties are stored in the info field and can be displayed by print function.

```{r}
db <- rocker::newDB() # New database handling object
db$setupPostgreSQL()
db$info
```

```{r}
db
```

```{r}
db$print()
```

```{r}
print(db)
```

```{r}
db$unloadDriver()
```

