# Additional packages and database types

The listed packages are required for some functions of *rocker*.

## *RSQLite* package

[*RSQLite*](https://github.com/r-dbi/RSQLite) package for handling of SQLite database connections.
It is required for the setupSQLite() function of *rocker* class.

```{r, eval = FALSE}
install.packages("RSQLite")
```

__Setup database__

Option 1

```{r}
db <- rocker::newDB() # New database handling object
db$setupSQLite( # Setup SQLite database
  dbname = ":memory:"
)
db$unloadDriver() # Reset database handling object
```

Option 2

```{r}
db <- rocker::newDB() # New database handling object
db$setupDriver( # Setup SQLite database
  drv = RSQLite::SQLite(),
  dbname = ":memory:"
)
db$unloadDriver() # Reset database handling object
```

## *RPostgres* package

[*RPostgres*](https://github.com/r-dbi/RPostgres) package for handling of PostgreSQL database connections.
It is required for the setupPostgreSQL() function of *rocker* class.

```{r, eval = FALSE}
install.packages("RPostgres")
```

__Setup database__

Option 1

```{r, eval = FALSE}
db <- rocker::newDB() # New database handling object
#> dctr | New object
db$setupPostgreSQL( # Setup PostgreSQL database
  host = "127.0.0.1", port = "5432", dbname = "mydb",
  user = "postgres", password = "password"
)
#> Dctr | Driver load RPostgres
db$unloadDriver() # Reset database handling object
#> dctr | Driver unload RPostgres
```

Option 2

```{r, eval = FALSE}
db <- rocker::newDB() # New database handling object
#> dctr | New object
db$setupDriver( # Setup PostgreSQL database
  drv = RPostgres::Postgres(),
  host = "127.0.0.1", port = "5432", dbname = "mydb",
  user = "postgres", password = "password"
)
#> Dctr | Driver load RPostgres
db$unloadDriver() # Reset database handling object
#> dctr | Driver unload RPostgres
```

## *RMariaDB* package

[*RMariaDB*](https://github.com/r-dbi/RMariaDB) package for handling of MariaDB and MySQL database connections.
It is required for the setupMariaDB() function of *rocker* class.

```{r, eval = FALSE}
install.packages("RMariaDB")
```

__Setup database__

Option 1

```{r, eval = FALSE}
db <- rocker::newDB() # New database handling object
#> dctr | New object
db$setupMariaDB( # Setup MariaDB database
  host = "127.0.0.1", port = "3306", dbname = "mydb",
  user = "root", password = "password"
)
#> Dctr | Driver load RMariaDB
db$unloadDriver() # Reset database handling object
#> dctr | Driver unload RMariaDB
```

Option 2

```{r, eval = FALSE}
db <- rocker::newDB() # New database handling object
#> dctr | New object
db$setupDriver( # Setup MariaDB database
  drv = RMariaDB::MariaDB(),
  host = "127.0.0.1", port = "3306", dbname = "mydb",
  user = "root", password = "password"
)
#> Dctr | Driver load RMariaDB
db$unloadDriver() # Reset database handling object
#> dctr | Driver unload RMariaDB
```

## *crayon* package

The [*crayon*](https://github.com/r-lib/crayon) package is required for colored terminal output.
If missing terminal output is monochrome.

```{r, eval = FALSE}
install.packages("crayon")
```
